---
title: "Trees"
format: html
---

```{r include=FALSE}
#package loading
library(tidyverse)
library(janitor)
library(ROCR)
library(tree)
library(randomForest)
```

## 1. Data Loading and Cleaning

```{r}
bank <- read_delim("../data/bank-additional-full.csv", delim = ";", show_col_types = FALSE) |>
  janitor::clean_names() |>
  mutate_if(is.character, as.factor) 
```

## 2. Train-Test Split

```{r}
set.seed(42)
split <- sample(seq_len(nrow(bank)), size = floor(0.8 * nrow(bank)))
bankTrain <- bank[split, ]
bankTest  <- bank[-split, ]
```

## 3. Paper-Style Decision Tree

This tree uses primarily the economic predictors available to us in the data that best imitates the predictors used in the paper.

```{r}
bank_tree_paper <- tree(y ~ housing + loan + default + month + emp_var_rate +
                          cons_price_idx + cons_conf_idx + euribor3m + nr_employed,
                        data = bankTrain)
summary(bank_tree_paper)
```

```{r}
plot(bank_tree_paper)
text(bank_tree_paper, pretty = 0)
```

Due to the small amount of $yes$ observations present in the data the tree algorithm decides the most optimal way to split is classifying every observation as $no$.

## 4. Extending the tree Model

This model seeks to include more variables to give better split options and hopefully create a tree that has at least a node that classifies an observation as $yes$.

```{r}
<<<<<<< HEAD
bank_tree_extended <- tree(y ~ job + marital + education + housing + loan +
                          month + day_of_week + campaign + emp_var_rate +
                          cons_price_idx + cons_conf_idx + euribor3m + nr_employed,
=======
bank_tree_extended <- tree(y ~ .-duration,
>>>>>>> origin/main
                        data = bankTrain)
summary(bank_tree_extended)
```

<<<<<<< HEAD
Based on the summary alone, this is still creating the same tree as before. To lesson the impact of $month$ and $nr\_employed$ we will use a Random Forest.

### Random Forest Model

```{r}
set.seed(42)
bank_rf <- randomForest(y ~ job + marital + education + housing + loan +
                          month + day_of_week + campaign + emp_var_rate +
                          cons_price_idx + cons_conf_idx + euribor3m + nr_employed,
=======
```{r}
plot(bank_tree_extended)
text(bank_tree_extended, pretty = 0)
```

The extended tree using all predictors, minus $duration$, does improve over the baseline tree. However, it is still using very few variables. We will attempt to use a random forest to draw on more variables in the dataset.

## 5. Random Forest Model

```{r}
set.seed(42)
bank_rf <- randomForest(y ~ .-duration,
>>>>>>> origin/main
                        data = bankTrain, importance = TRUE)
bank_rf
```

```{r}
importance(bank_rf)
```

<<<<<<< HEAD
From these results we can see a definite improvement over the previous trees. The Out-Of-Bag error rate is at $10.75\%$ indicating that there are trees in the random forest classifying observations as $yes$. This error rate is also lower than rate seen in the previous trees. Though it is important to note the high class error rate for $yes$ of $0.75$. The Random Forest has also decided on different variables as being important to the classification. It has placed $nr\_employed$ as the least important and $job$, $marital$, $education$, $housing$, and $loan$ all before $month$.

#### Test prediction of random forest

```{r}
rf_preds <- predict(bank_rf, newdata = bankTest, type = "response")
mean(rf_preds != bankTest$y)
```

The Out-Of-Bag estimation of error was very close for the random forest. As the test misclassification rate was $10.8\%$.
=======
From these results we can see an improvement over the previous trees. The Out-Of-Bag error rate is at $10.19\%$. Though it is important to note the high class error rate for $yes$ of $0.72$. The Random Forest has greatly expanded the variables used in the classification process, $nr\_employed$ has been determined to be the least important variable in the classification process.

#### AUC of Random Forest

```{r}
rf_auc_preds <- predict(bank_rf, newdata = bankTest, type = "prob")[,2]
roc.predictions <- prediction(rf_auc_preds, bankTest$y)
roc.perf <- performance(roc.predictions, "tpr", "fpr")
plot(roc.perf, col = "blue", lwd = 2, main = "ROC Curve – Random Forest")
abline(a = 0, b = 1, col = "black", lty = 2)
rf_auc <- performance(roc.predictions, "auc")@y.values[[1]]
rf_auc
```

The AUC of the tree model results in $0.802$.

### Test Error Rate of Random Forest and Extended Tree

```{r}
set.seed(42)
rf_preds <- predict(bank_rf, newdata = bankTest, type = "class")
mean(rf_preds != bankTest$y)
```

```{r}
extended_preds <- predict(bank_tree_extended, newdata = bankTest, type = "class")
mean(extended_preds != bankTest$y)
```

Interestingly the test misclassification error rate for the two trees is almost identical with the extended tree slightly edging out the random forest.

## Tuning the Random Forest

```{r}
#Get design matrix
bank_design <- model.matrix(bank_tree_extended)[,-1]

set.seed(42)
tuneRF(bank_design, bankTrain$y, mtryStart = 4, ntreeTry = 500, stepFactor = 1, plot = FALSE)
```

The tuning function is not showing any improvement over the default used in the original model, so we will leave it as is
>>>>>>> origin/main
